---
title: Large Bayes Factors, but Zero in CrI
subtitle: Minimal Working Example
output: html_document
date: "`r Sys.Date()`"
author: Michael Vrazitulis
---

In this document, I show the puzzling issue that I encountered when analyzing my master's thesis data.
In short, there are large Bayes factors (up to $BF_{10} = 18.8$) in cases where the 95% CrI of the parameter of interest includes zero.

For starters, let's load `ggplot2` and `brms`:

```{r}
library(ggplot2)
library(brms)
```

Now we can load in the data set of interest. I've stripped it of anything but the columns necessary for downstream analysis.

(If you don't have the TSV file at hand, here's a download link: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YMcP87g5RcUY-dYg?e=6vb4Me>, 235.9 KiB)

```{r}
critical_results_SPV <- read.table(file = "critical_results_SPV.tsv", sep = "\t", header = TRUE)
```

How does this data look like? Let's find out.

```{r}
str(critical_results_SPV)
summary(critical_results_SPV)
```

Clarifications:

The **subject** column contains IDs of the particular subjects.

The **item** column contains IDs of items. This experimental design has only four items.

The **VR** column contains the Good/Bad response provided by the subject.  
Here, Good is sum-coded as 0.412 and Bad is as --0.588.

The **POL** column contains information on whether we are in positive- or negative-polarity condition.  
Here, positive is sum-coded as 0.498 and negative as --0.502.

Finally, the **res_log_RT** column contains our dependent variable of interest (residual log RT).  
It was derived by first log-transforming raw RTs on the millisecond scale, then subtracting predicted log RTs produced by a model that had been fitted separately on control-condition trials.  
(In hindsight, I realize that this type of residualization preprocessing is kind of cringe, but I didn't know that a year ago. Now it is what it is.)  

As a sanity check, let's look at how the dependent variable is distributed (histogram):

```{r}
hist(critical_results_SPV$res_log_RT)
```

At this point, we are ready to fit some `brms` models.

The hypothesis we want to test is that the interaction of VR (verification response) and POL (polarity) has a meaningful effect on residual log RTs. According to theory, we should expect a negative parameter value $\beta_{\mathtt{\,VR:POL}}$.

(Disclaimer: In order to ensure convergence, I fitted each of the models with 40k iterations! That takes forever, though, so I have set the code blocks below where models are fitted to `eval=FALSE`. But I provide a download link of an .RData file for each of the fitted models. So, in case you want to reproduce any downstream analyses *after* models have been fitted, feel free to just download my fitted models and load them into your workspace.)  
&nbsp;

```{r eval=FALSE}
## null model (no population-level VR:POL), for comparison using Bayes factors

priors_null <-  c(
  prior(normal(0, 1), class = Intercept), # 0 because residual log RT
  prior(normal(0, 0.3), class = sigma),
  prior(normal(0, 0.2), class = b, coef = VR),
  prior(normal(0, 0.2), class = b, coef = POL),
  prior(normal(0, 0.1), class = sd),
  prior(lkj(2), class = cor)
)

fit_null <- brm('res_log_RT ~ VR + POL + (VR * POL | subject) + (VR * POL | item)',
                data = critical_results_SPV,
                family = gaussian(),
                prior = fit_null,
                chains = 4,
                cores = 12,
                iter = 40000,
                warmup = 8000,
                control = list(adapt_delta = 0.95),
                save_pars = save_pars(all = TRUE)
)
```

Download link for `fit_null`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YT6nAr7Y05C91zdA> (3.8 GiB)  
&nbsp;

```{r eval=FALSE}
## target parameter with prior N(0, 1)

priors_interest_prior_1 <- c(
  prior(normal(0, 1), class = Intercept), # 0 because residual log RT
  prior(normal(0, 0.3), class = sigma),
  prior(normal(0, 0.2), class = b, coef = VR),
  prior(normal(0, 0.2), class = b, coef = POL),
  prior(normal(0, 1), class = b, coef = VR:POL), # target!
  prior(normal(0, 0.1), class = sd),
  prior(lkj(2), class = cor)
)

fit_interest_prior_1 <- brm(res_log_RT ~ VR * POL + (VR * POL | subject) + (VR * POL | item),
                            data = critical_results_SPV,
                            family = gaussian(),
                            prior = priors_interest_prior_1,
                            chains = 4,
                            cores = 12,
                            iter = 40000,
                            warmup = 8000,
                            control = list(adapt_delta = .95),
                            save_pars = save_pars(all = TRUE)
)
```

Download link for `fit_interest_prior_1`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YVZdXjZ4MsyceBqQ?e=zRs6AM> (3.8 GiB)  
&nbsp;

```{r eval=FALSE}
## target parameter with prior N(0, 0.5)

priors_interest_prior_0_5 <- c(
  prior(normal(0, 1), class = Intercept), # 0 because residual log RT
  prior(normal(0, 0.3), class = sigma),
  prior(normal(0, 0.2), class = b, coef = VR),
  prior(normal(0, 0.2), class = b, coef = POL),
  prior(normal(0, 0.5), class = b, coef = VR:POL), # target!
  prior(normal(0, 0.1), class = sd),
  prior(lkj(2), class = cor)
)

fit_interest_prior_0_5 <- brm(res_log_RT ~ VR * POL + (VR * POL | subject) + (VR * POL | item),
                              data = critical_results_SPV,
                              family = gaussian(),
                              prior = priors_interest_prior_0_5,
                              chains = 4,
                              cores = 12,
                              iter = 40000,
                              warmup = 8000,
                              control = list(adapt_delta = .95),
                              save_pars = save_pars(all = TRUE)
)
```

Download link for `fit_interest_prior_0_5`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YQnHW07js2r_lXHQ?e=rYWPCT> (3.8 GiB)  
&nbsp;

```{r eval=FALSE}
## target parameter with prior N(0, 0.2)

priors_interest_prior_0_2 <- c(
  prior(normal(0, 1), class = Intercept), # 0 because residual log RT
  prior(normal(0, 0.3), class = sigma),
  prior(normal(0, 0.2), class = b, coef = VR),
  prior(normal(0, 0.2), class = b, coef = POL),
  prior(normal(0, 0.2), class = b, coef = VR:POL), # target!
  prior(normal(0, 0.1), class = sd),
  prior(lkj(2), class = cor)
)

fit_interest_prior_0_2 <- brm(res_log_RT ~ VR * POL + (VR * POL | subject) + (VR * POL | item),
                              data = critical_results_SPV,
                              family = gaussian(),
                              prior = priors_interest_prior_0_2,
                              chains = 4,
                              cores = 12,
                              iter = 40000,
                              warmup = 8000,
                              control = list(adapt_delta = .95),
                              save_pars = save_pars(all = TRUE)
)
```

Download link for `fit_interest_prior_0_2`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YPn7AuIWpVbe-pLQ?e=tgbxkt> (3.8 GiB)  
&nbsp;

```{r eval=FALSE}
## target parameter with prior N(0, 0.1)

priors_interest_prior_0_1 <- c(
  prior(normal(0, 1), class = Intercept), # 0 because residual log RT
  prior(normal(0, 0.3), class = sigma),
  prior(normal(0, 0.2), class = b, coef = VR),
  prior(normal(0, 0.2), class = b, coef = POL),
  prior(normal(0, 0.1), class = b, coef = VR:POL), # target!
  prior(normal(0, 0.1), class = sd),
  prior(lkj(2), class = cor)
)

fit_interest_prior_0_1 <- brm(res_log_RT ~ VR * POL + (VR * POL | subject) + (VR * POL | item),
                              data = critical_results_SPV,
                              family = gaussian(),
                              prior = priors_interest_prior_0_1,
                              chains = 4,
                              cores = 12,
                              iter = 40000,
                              warmup = 8000,
                              control = list(adapt_delta = .95),
                              save_pars = save_pars(all = TRUE)
)
```

Download link for `fit_interest_prior_0_1`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YSqygFgrJt9wD5rw?e=nxVJEr> (3.8 GiB)  
&nbsp;

```{r eval=FALSE}
## target parameter with prior N(0, 0.05)

priors_interest_prior_0_05 <- c(
  prior(normal(0, 1), class = Intercept), # 0 because residual log RT
  prior(normal(0, 0.3), class = sigma),
  prior(normal(0, 0.2), class = b, coef = VR),
  prior(normal(0, 0.2), class = b, coef = POL),
  prior(normal(0, 0.05), class = b, coef = VR:POL), # target!
  prior(normal(0, 0.1), class = sd),
  prior(lkj(2), class = cor)
)

fit_interest_prior_0_05 <- brm(res_log_RT ~ VR * POL + (VR * POL | subject) + (VR * POL | item),
                               data = critical_results_SPV,
                               family = gaussian(),
                               prior = priors_interest_prior_0_05,
                               chains = 4,
                               cores = 12,
                               iter = 40000,
                               warmup = 8000,
                               control = list(adapt_delta = .95),
                               save_pars = save_pars(all = TRUE)
)
```

Download link for `fit_interest_prior_0_05`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YUMqrpeMfDth4otw?e=pOvwKQ> (3.8 GiB)  
&nbsp;

```{r eval=FALSE}
## target parameter with prior N(0, 0.02)

priors_interest_prior_0_02 <- c(
  prior(normal(0, 1), class = Intercept), # 0 because residual log RT
  prior(normal(0, 0.3), class = sigma),
  prior(normal(0, 0.2), class = b, coef = VR),
  prior(normal(0, 0.2), class = b, coef = POL),
  prior(normal(0, 0.02), class = b, coef = VR:POL), # target!
  prior(normal(0, 0.1), class = sd),
  prior(lkj(2), class = cor)
)

fit_interest_prior_0_02 <- brm(res_log_RT ~ VR * POL + (VR * POL | subject) + (VR * POL | item),
                               data = critical_results_SPV,
                               family = gaussian(),
                               prior = priors_interest_prior_0_02,
                               chains = 4,
                               cores = 12,
                               iter = 40000,
                               warmup = 8000,
                               control = list(adapt_delta = .95),
                               save_pars = save_pars(all = TRUE)
)
```

Download link for `fit_interest_prior_0_02`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YWdPcqTiPKRiSiBA?e=y7GfNW> (3.8 GiB)  
&nbsp;

```{r eval=FALSE}
## target parameter with prior N(0, 0.01)

priors_interest_prior_0_01 <- c(
  prior(normal(0, 1), class = Intercept), # 0 because residual log RT
  prior(normal(0, 0.3), class = sigma),
  prior(normal(0, 0.2), class = b, coef = VR),
  prior(normal(0, 0.2), class = b, coef = POL),
  prior(normal(0, 0.01), class = b, coef = VR:POL), # target!
  prior(normal(0, 0.1), class = sd),
  prior(lkj(2), class = cor)
)

fit_interest_prior_0_01 <- brm(res_log_RT ~ VR * POL + (VR * POL | subject) + (VR * POL | item),
                               data = critical_results_SPV,
                               family = gaussian(),
                               prior = priors_interest_prior_0_01,
                               chains = 4,
                               cores = 12,
                               iter = 40000,
                               warmup = 8000,
                               control = list(adapt_delta = .95),
                               save_pars = save_pars(all = TRUE)
)
```

Download link for `fit_interest_prior_0_01`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YR37krhRDvYPaNhg?e=c6wSLq> (3.8 GiB)  
&nbsp;

Next, we can compute marginal likelihoods. This is a little time-consuming (but not quite as bad as model fitting). So I have done that in advance as well. Download links are provided for each marginal-likelihood object (stored as .RData files). The code blocks below where the bridge-sampling happens are set to `eval=FALSE` here.

```{r eval=FALSE}
# null model (constrains VR:POL parameter to zero)
load("fit_null.RData") # to load in downloaded file
set.seed(1)
margLogLik_null <- bridge_sampler(fit_null, maxiter=10000)
rm(fit_null)
```

Download link for `margLogLik_null`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9Ya_1SCv-XeSNoShA?e=ungExU> (4.9 MiB)  
&nbsp;
```{r eval=FALSE}
# model with prior N(0, 1) on VR:POL parameter
load("fit_interest_prior_1.RData") # to load in downloaded file
set.seed(1)
margLogLik_interest_prior_1 <- bridge_sampler(fit_interest_prior_1, maxiter=10000)
rm(fit_interest_prior_1)
```

Download link for `margLogLik_interest_prior_1`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YYASBtuzqJMIhsng?e=zRX8h7> (5.6 MiB)  
&nbsp;

```{r eval=FALSE}
# model with prior N(0, 0.5) on VR:POL parameter
load("fit_interest_prior_0_5.RData") # to load in downloaded file
set.seed(1)
margLogLik_interest_prior_0_5 <- bridge_sampler(fit_interest_prior_0_5, maxiter=10000)
rm(fit_interest_prior_0_5)
```

Download link for `margLogLik_interest_prior_0_5`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9Yey12ZOVnjNVxP4Q?e=iHBOb8> (5.7 MiB)  
&nbsp;

```{r eval=FALSE}
# model with prior N(0, 0.2) on VR:POL parameter
load("fit_interest_prior_0_2.RData") # to load in downloaded file
set.seed(1)
margLogLik_interest_prior_0_2 <- bridge_sampler(fit_interest_prior_0_2, maxiter=10000)
rm(fit_interest_prior_0_2)
```

Download link for `margLogLik_interest_prior_0_2`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YdpMVpLFz0_Aw-qA?e=jDDvFh> (5.5 MiB)  
&nbsp;

```{r eval=FALSE}
# model with prior N(0, 0.1) on VR:POL parameter
load("fit_interest_prior_0_1.RData") # to load in downloaded file
set.seed(1)
margLogLik_interest_prior_0_1 <- bridge_sampler(fit_interest_prior_0_1, maxiter=10000)
rm(fit_interest_prior_0_1)
```

Download link for `margLogLik_interest_prior_0_1`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YXbgo3xb3q9YKQVw?e=BDqpOK> (4.4 MiB)  
&nbsp;

```{r eval=FALSE}
# model with prior N(0, 0.05) on VR:POL parameter
load("fit_interest_prior_0_05.RData") # to load in downloaded file
set.seed(1)
margLogLik_interest_prior_0_05 <- bridge_sampler(fit_interest_prior_0_05, maxiter=10000)
rm(fit_interest_prior_0_05)
```

Download link for `margLogLik_interest_prior_0_05`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YUMqrpeMfDth4otw?e=8VzKkB> (4.7 MiB)  
&nbsp;

```{r eval=FALSE}
# model with prior N(0, 0.02) on VR:POL parameter
load("fit_interest_prior_0_02.RData") # to load in downloaded file
set.seed(1)
margLogLik_interest_prior_0_02 <- bridge_sampler(fit_interest_prior_0_02, maxiter=10000)
rm(fit_interest_prior_0_02)
```

Download link for `margLogLik_interest_prior_0_02`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9Ybvd8xI-3vKnVaEg?e=LE8bbs> (5.0 MiB)  
&nbsp;

```{r eval=FALSE}
# model with prior N(0, 0.01) on VR:POL parameter
load("fit_interest_prior_0_01.RData") # to load in downloaded file
set.seed(1)
margLogLik_interest_prior_0_01 <- bridge_sampler(fit_interest_prior_0_01, maxiter=10000)
rm(fit_interest_prior_0_01)
```

Download link for `margLogLik_interest_prior_0_01`: <https://1drv.ms/u/s!AuRxgm6Ml5A4l9YZ66iQ0gBqPshwhA?e=Ye7ImF> (5.2 MiB)  
&nbsp;

Finally, we can calculate Bayes factors:

```{r}
load("margLogLik_null.RData") # to load in downloaded file
```

```{r}
# BF10 for model with VR:POL prior N(0, 1) over null model
load("margLogLik_interest_prior_1.RData") # to load in downloaded file
BF_prior_1 <- bayes_factor(margLogLik_interest_prior_1, margLogLik_null)
BF_prior_1$bf

# BF10 for model with VR:POL prior N(0, 0.5) over null model
load("margLogLik_interest_prior_0_5.RData") # to load in downloaded file
BF_prior_0_5 <- bayes_factor(margLogLik_interest_prior_0_5, margLogLik_null)
BF_prior_0_5$bf

# BF10 for model with VR:POL prior N(0, 0.2) over null model
load("margLogLik_interest_prior_0_2.RData") # to load in downloaded file
BF_prior_0_2 <- bayes_factor(margLogLik_interest_prior_0_2, margLogLik_null)
BF_prior_0_2$bf

# BF10 for model with VR:POL prior N(0, 0.1) over null model
load("margLogLik_interest_prior_0_1.RData") # to load in downloaded file
BF_prior_0_1 <- bayes_factor(margLogLik_interest_prior_0_1, margLogLik_null)
BF_prior_0_1$bf

# BF10 for model with VR:POL prior N(0, 0.05) over null model
load("margLogLik_interest_prior_0_05.RData") # to load in downloaded file
BF_prior_0_05 <- bayes_factor(margLogLik_interest_prior_0_05, margLogLik_null)
BF_prior_0_05$bf

# BF10 for model with VR:POL prior N(0, 0.02) over null model
load("margLogLik_interest_prior_0_02.RData") # to load in downloaded file
BF_prior_0_02 <- bayes_factor(margLogLik_interest_prior_0_02, margLogLik_null)
BF_prior_0_02$bf

# BF10 for model with VR:POL prior N(0, 0.01) over null model
load("margLogLik_interest_prior_0_01.RData") # to load in downloaded file
BF_prior_0_01 <- bayes_factor(margLogLik_interest_prior_0_01, margLogLik_null)
BF_prior_0_01$bf
```

As we can see, the $BF_{10}$ ends up being $> 10$ in two cases, seemingly indicating strong evidence for the alternative:

- $BF_{10} = 18.8$, in case of model with prior N(0, 0.1),
- $BF_{10} = 12.2$, in case of model with prior N(0, 0.05).

But---here comes the bizarre thing---in both of these fitted models, the 95% CrI of VR:POL comfortably includes zero:

```{r}
# population-level effects for model with prior N(0, 0.1) where e had BF10 = 12.2
load("fit_interest_prior_0_1.RData") # to load in downloaded file
fixef(fit_interest_prior_0_1)

# population-level effects for model with prior N(0, 0.05) where e had BF10 = 18.8
load("fit_interest_prior_0_05.RData") # to load in downloaded file
fixef(fit_interest_prior_0_05)
```

Lastly, for a complete picture, we can plot the prior and posterior distributions for the case with prior N(0, 0.05), i.e., with Bayes factor 12.2:

```{r}
set.seed(2)
posterior_samples <- as_draws_df(fit_interest_prior_0_05)
prior_samples <- data.frame(b = rnorm(n = nrow(posterior_samples), mean = 0, sd = 0.05))
colnames(prior_samples) <- "b_VR:POL"
combined_samples <- rbind(
  data.frame(Value = prior_samples$`b_VR:POL`, Distribution = "Prior"),
  data.frame(Value = posterior_samples$`b_VR:POL`, Distribution = "Posterior")
)
ggplot(combined_samples, aes(x = Value, fill = Distribution)) +
  geom_density(alpha = 0.5) +
  labs(title = "Prior N(0, 0.05) and Posterior distribution of b_VR:POL, where BF10 = 12.2",
       x = "b_VR:POL",
       y = "Density") +
  scale_fill_manual(values = c("red", "blue")) +
  theme_minimal()
```
